---
title: "DISSECT"
author: "Sergej Ruff"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{DISSECT}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

DISSECT trains an ensemble with semi-supervised consistency regularization: at
each step it forms mixtures of simulated and real samples, and penalises
predictions on the mixture that disagree with the corresponding mixture of the
individual predictions. Because that constraint uses the real samples without
needing their labels, the model sees target-domain data during training.

It estimates cell-type fractions and, optionally, cell-type-specific
expression.

Chunks below are not evaluated during build. Step counts are reduced for
demonstration; use the function defaults for real analyses.

# Data

```{r data}
library(quasar)
library(QuasarDeconData)

sc_ref <- load_cov_imm("train")
pbulk  <- load_cov_pbulk(1)

bulk  <- pbulk$bulk_expression_profiles   # genes x samples
truth <- pbulk$ground_truth_proportions
```

# Simulate

```{r simulate}
dissect_sim <- dissect_simulate(
  sc_data      = sc_ref,
  celltype_col = "cell_type",
  batch_col    = "donor_id",
  type         = "bulk",
  n_samples    = 5000,
  save_expr    = TRUE,
  seed         = 42
)
```

`save_expr = TRUE` stores per-cell-type expression contributions for every
simulated sample. It increases memory use considerably but is a hard
requirement for the expression model — `dissect_expr()` errors without it. If
you only want proportions, set it to `FALSE`.

`batch_col` performs simulation separately per batch and concatenates the
results, so batch structure in the reference is preserved rather than mixed
away.

Setting `type = "st"` produces spot-level profiles instead: only a handful of
cells per spot, allocated by rounding rather than flooring. `downsample` then
thins the resulting counts without replacement.

# Process

```{r process}
dissect_proc <- dissect_process(
  bulk        = bulk,             # genes x samples
  sim_data    = dissect_sim,
  var_cutoff  = 0.1,
  test_in_mix = 1
)
```

`test_in_mix` controls how many real samples enter the online mixing step. For
microarray data, pass `test_dataset_type = "microarray"` — the values are
back-transformed out of log space before normalisation.

# Estimate fractions

```{r prop}
dissect_prop_res <- dissect_prop(
  processed = dissect_proc,
  n_steps   = 5000,
  models    = 1:5,                # ensemble size
  device    = "auto"
)

head(dissect_prop_res$fractions)
head(dissect_prop_res$scores)
```

The consistency term is introduced in stages: the first 2000 steps use the
supervised loss alone, after which the consistency weight switches on. `scores`
holds the pre-activation outputs, before the softmax.

# Estimate expression

```{r expr}
dissect_expr_res <- dissect_expr(
  bulk         = bulk,
  fractions    = dissect_prop_res$fractions,
  sim_data     = dissect_sim,
  n_steps_expr = 5000,
  device       = "auto"
)

names(dissect_expr_res$expression_layered)
dim(dissect_expr_res$expression_layered[[1]])
```

Expression estimation is a separate stage that takes fractions as input, via a
conditional variational autoencoder. If you want to evaluate the expression
model in isolation, supply **ground-truth** proportions here instead of
estimated ones, so that expression differences are not confounded by errors in
the fractions.

`expression_layered` is one samples × genes matrix per cell type;
`expression_combined` is the same content as a long data frame with `cell_type`
and `sample` columns.

# Wrapper

```{r wrapper}
dissect_res <- dissect(
  sc_data      = sc_ref,
  bulk         = bulk,
  celltype_col = "cell_type",
  batch_col    = "donor_id",
  save_expr    = TRUE,
  device       = "auto"
)

head(dissect_res$fractions)
```

The wrapper runs simulation, processing, fraction estimation, and — when
`save_expr = TRUE` — expression estimation. If the expression stage fails it
warns and returns the proportion results rather than aborting.

# Evaluate

```{r evaluate}
quasar_prop_metrics(dissect_prop_res$fractions, truth)$cell_type_rmse
```

# Reference

Khatri, R., Machart, P., & Bonn, S. (2024). DISSECT: deep semi-supervised
consistency regularization for accurate cell type fraction and gene expression
estimation. *Genome Biology*, 25(1), 112.

If you use this implementation, cite both the QUASAR paper and Khatri et al.
