---
title: "OmicsTweezer"
author: "Sergej Ruff"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{OmicsTweezer}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

OmicsTweezer adds a domain-adaptation term to the training objective,
encouraging the encoder to place simulated pseudo-bulks and real bulk samples
in a shared embedding space. Each training batch pairs labelled source samples
with unlabelled target samples, so the alignment is learned without needing
target labels.

Chunks below are not evaluated during build. Epochs are reduced for
demonstration; use the function defaults for real analyses.

# Data

```{r data}
library(quasar)
library(QuasarDeconData)

sc_ref <- load_cov_imm("train")
pbulk  <- load_cov_pbulk(1)

bulk  <- pbulk$bulk_expression_profiles   # genes x samples
truth <- pbulk$ground_truth_proportions
```

# Simulate

```{r simulate}
ot_sim <- omics_simulate(
  sc_data      = sc_ref,
  celltype_col = "cell_type",
  samplenum    = 5000,
  n            = 500,
  random_state = 42
)
```

`sparse = TRUE` and `rare = TRUE` add optional perturbations: the first zeroes
a subset of cell types in a subset of samples, the second forces selected types
to very small fractions to emulate rare populations.

# Process

```{r process}
ot_proc <- omics_process(
  simudata           = ot_sim,
  real_bulk          = bulk,      # genes x samples
  variance_threshold = 0.98,
  scaler             = "ss"
)
```

`scaler` is `"ss"` for per-sample standardisation, `"mms"` for min-max, or
`"none"`.

# Train

```{r train}
ot_model <- omics_train(
  train_x       = ot_proc$train_x,
  train_y       = ot_proc$train_y,
  test_x        = ot_proc$test_x,
  dims          = c(512L, 256L, 128L, 64L),
  drops         = c(0, 0.3, 0.2, 0.1),
  epochs        = 30,
  batch_size    = 128,
  learning_rate = 1e-4,
  loss_weight   = 1.0,
  device        = "auto"
)

head(ot_model$history)
```

Note that `test_x` is passed to the *training* function. It supplies the
unlabelled target-domain batches for the adaptation term; no labels are used.

`loss_weight` controls how strongly the domain-adaptation term contributes
relative to the supervised MSE. Raising it pushes harder toward alignment at
the cost of supervised accuracy on the simulated data.

`dims` and `drops` must each be length four, matching the encoder and predictor
block structure.

# Predict

```{r predict}
ot_pred <- omics_predict(
  model      = ot_model,
  test_x     = ot_proc$test_x,
  celltypes  = ot_proc$celltypes,
  samplename = ot_proc$samplename,
  device     = "auto"
)

head(ot_pred)
```

# Wrapper

```{r wrapper}
ot_res <- omics_tweezer(
  sc_data      = sc_ref,
  real_bulk    = bulk,
  celltype_col = "cell_type",
  samplenum    = 5000,
  epochs       = 30,
  scaler       = "ss",
  device       = "auto"
)

head(ot_res$pred)
names(ot_res$per_model)
head(ot_res$loss_history)
```

The wrapper trains three architectures (`m256`, `m512`, `m1024`) and averages
their predictions. `per_model` keeps the individual predictions and
`loss_history` the epoch-level losses for all three.

`scale_minmax = TRUE` applies an additional global min-max scaling to the raw
count matrices before processing, mirroring the original `train_predict(scale =
TRUE)` path.

# Evaluate

```{r evaluate}
quasar_prop_metrics(ot_res$pred, truth)$cell_type_rmse
```

# Reference

Yang, X., Zhao, F., Ren, T., Chen, C., Byrne, K. T., Danilov, A. V., ... &
Xia, Z. (2025). OmicsTweezer: A distribution-independent cell deconvolution
model for multi-omics data. *Cell Genomics*, 5(9).

If you use this implementation, cite both the QUASAR paper and Yang et al.
