---
title: "QUASAR data"
author: "Sergej Ruff"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{QUASAR data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

Example and benchmark datasets live in a companion package, `QuasarDeconData`,
so that the main package stays light.

```{r install}
# install.packages("pak")
pak::pak("SergejRuff/QuasarDeconData")

library(QuasarDeconData)
```

Every deconvolution method in this package needs two inputs: an annotated
single-cell reference and a bulk expression matrix. The datasets below supply
matched pairs of both.

# COVID-19 PBMC

A labelled scRNA-seq dataset of PBMCs from individuals with SARS-CoV-2
infection, partitioned at the **donor** level into training and test halves
with disease severity balanced across both. Donor-level splitting means no cell
contributes to both the reference and the targets.

```{r cov-sc}
sc_train <- load_cov_imm("train")
sc_test  <- load_cov_imm("test")

sc_train
table(sc_train$cell_type)
table(sc_train$donor_id)
```

Cell-type labels are in `cell_type`, donor labels in `donor_id`. `donor_id` is
retained because some deconvolution methods require per-subject information,
and because donor-specific pseudo-bulk generation preserves expression
heterogeneity that pooling would average away.

A third reference contains only non-diseased donors:

```{r cov-healthy}
sc_healthy <- load_cov_imm("healthy")
```

This one exists for **reference-mismatch experiments** — deconvolving
COVID-19 mixtures with a reference built from healthy donors, which is the
realistic situation when condition-matched single-cell data is unavailable. It
may lack cell types present in the benchmark ground truth.

## Pseudo-bulk

Ten independently simulated test datasets were generated from the held-out test
cells, each with 1000 pseudo-bulk samples at 100 cells per sample.

```{r cov-pbulk}
pbulk <- load_cov_pbulk(1)          # a single test dataset

bulk  <- pbulk$bulk_expression_profiles   # genes x samples
truth <- pbulk$ground_truth_proportions   # samples x cell types

dim(bulk)
dim(truth)
head(truth)
```

```{r cov-pbulk-all}
all_pbulk <- load_cov_pbulk()        # all ten as a named list
train_pb  <- load_cov_pbulk("train")    # from the training split
healthy_pb <- load_cov_pbulk("healthy") # from the healthy-only split
```

The training pseudo-bulk contains 10000 samples at 500 cells each, and is what
every deep-learning method was given as common training data in the benchmark.
The train and healthy sets are stored as chunks and reassembled on load, so the
split is invisible to you and sample order is preserved.

# 10x Genomics PBMC

A reference combining the publicly available 6k, 8k, and 10k healthy-donor PBMC
datasets from 10x Genomics, each a different donor. 22,629 features across
21,978 cells.

```{r pbmc}
sc_pbmc <- load_pbmc_sc_ref()

table(sc_pbmc$celltype)
table(sc_pbmc$donor_id)
```

Note the cell-type column here is **`celltype`**, not `cell_type` as in the
COVID-19 data. Combining three donors increases reference size and supplies the
donor information that methods like MuSiC and MEAD require.

Matching pseudo-bulk:

```{r pbmc-pb}
pb <- load_pbmc_sc_ref("pseudobulk")

dim(pb$bulk_expression_profiles)
dim(pb$ground_truth_proportions)
```

The original 10x datasets are licensed CC BY 4.0. The versions here were
processed, annotated, filtered, and reformatted for QUASAR workflows; no
endorsement by 10x Genomics is implied.

# Real bulk benchmarks

Four PBMC datasets with flow-cytometry-derived ground-truth proportions. Use
these with `load_pbmc_sc_ref()` as the reference.

```{r real}
data(GSE107011)   # 12 samples,  36,128 features, RNA-seq
data(GSE107572)   #  9 samples,  19,423 features, RNA-seq
data(GSE120502)   # 249 samples, 39,376 features, RNA-seq
data(GSE65133)    # 20 samples,  35,688 features, microarray

names(GSE107011)
dim(GSE107011$bulk_expression_profiles)
colnames(GSE107011$ground_truth_proportions)
```

All four share the same six cell-type columns — `Bcells`, `CD4Tcells`,
`CD8Tcells`, `Monocytes`, `NK`, `Unknown` — though the column *order* differs
between datasets. `quasar_prop_metrics()` aligns by name, so order does not
matter when scoring.

`GSE65133` is microarray rather than RNA-seq. DISSECT handles this explicitly:

```{r microarray}
dissect_micro <- dissect(
  sc_data           = sc_pbmc,
  bulk              = GSE65133$bulk_expression_profiles,
  celltype_col      = "celltype",
  test_dataset_type = "microarray",
  device            = "auto"
)
```

Ground-truth proportions for these four follow the benchmark used by Khatri et
al. (2024); the bulk expression matrices were retrieved independently from GEO.

# Orientation summary

| Object | Layout |
|---|---|
| `bulk_expression_profiles` | genes × samples |
| `ground_truth_proportions` | samples × cell types |
| Single-cell references | `Seurat`, features × cells |

Most `*_process()` functions expect genes × samples, so the packaged bulk
matrices can be passed straight through.

# Sources

- COVID-19 Immune Atlas, CELLxGENE collection
  `b9fc3d70-5a72-4479-a046-c2cc1ab19efc`
- 10x Genomics 6k / 8k / 10k healthy-donor PBMC datasets (CC BY 4.0)
- GEO accessions GSE107011, GSE107572, GSE120502, GSE65133
