---
title: "Scaden"
author: "Sergej Ruff"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Scaden}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

Scaden trains an ensemble of three multilayer perceptrons of increasing width
(`m256`, `m512`, `m1024`) on simulated pseudo-bulks and averages their
predictions. It is the simplest of the four ported methods and a reasonable
first thing to run on a new dataset.

Chunks below are not evaluated during build. Sample counts and epochs are
reduced for demonstration; use the function defaults for real analyses.

# Data

```{r data}
library(quasar)
library(QuasarDeconData)

sc_ref <- load_cov_imm("train")
pbulk  <- load_cov_pbulk(1)

bulk  <- pbulk$bulk_expression_profiles   # genes x samples
truth <- pbulk$ground_truth_proportions   # samples x cell types
```

# Simulate

```{r simulate}
scaden_sim <- scaden_sim_pb(
  sc_data      = sc_ref,
  celltype_col = "cell_type",
  n            = 500,      # cells per pseudo-bulk
  samplenum    = 5000,     # number of pseudo-bulks
  sparse       = TRUE,
  seed         = 42
)
```

`sparse = TRUE` zeroes a subset of cell types in a subset of samples, producing
training mixtures in which some populations are absent. This matters when the
target samples may not contain every cell type in the reference.

`scaden_sim_pb()` can also merge several cell types into a single unknown
class via `unknown_celltypes`. The merge happens *after* simulation — each
original type gets its own Dirichlet component and is sampled independently,
then the proportion columns are summed. This mirrors the original Python
behaviour and differs from relabelling cells before simulation.

# Process

```{r process}
scaden_proc <- scaden_process(
  sim_data   = scaden_sim,
  bulk_data  = bulk,
  mode       = "scaden",
  var_cutoff = 0.1
)
```

Two preprocessing modes are available:

- `"scaden"` applies an absolute variance cutoff to the bulk matrix only, then
  intersects with the training genes. Uses `log2(x + 1)` and per-sample min-max
  scaling.
- `"tape"` applies a top-quantile cutoff to both matrices independently before
  intersecting. Uses `log(x + 1)` and either min-max or standardisation via
  `scaler`.

The log base is immaterial in practice — both scalers are invariant to
multiplication by a positive constant — but each mode stays literal to its
source.

# Train

```{r train}
scaden_fit <- scaden(
  train_x    = scaden_proc$train_x,
  train_y    = scaden_proc$train_y,
  lr         = 1e-4,
  batch_size = 128,
  epochs     = 20,
  seed       = 123
)
```

All three architectures train sequentially, each with its own progress bar.
Per-batch and per-epoch losses are returned for every model.

# Predict

```{r predict}
scaden_pred <- scaden_predict(scaden_fit, scaden_proc$test_x)

head(scaden_pred$average_output)
```

The individual ensemble members are returned as well
(`prediction_model256`, `prediction_model512`, `prediction_model1024`), which
is useful for gauging how much the three architectures disagree — wide
disagreement on a sample is a signal worth following up.

# Evaluate

```{r evaluate}
metrics <- quasar_prop_metrics(scaden_pred$average_output, truth)

metrics$cell_type_rmse
metrics$pearson_celltype_cor
```

# Reference

Menden, K., Marouf, M., Oller, S., Dalmia, A., Magruder, D. S., Kloiber, K.,
... & Bonn, S. (2020). Deep learning-based cell composition analysis from
tissue expression profiles. *Science Advances*, 6(30), eaba2619.

If you use this implementation, cite both the QUASAR paper and Menden et al.
