---
title: "TAPE"
author: "Sergej Ruff"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{TAPE}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

TAPE uses an autoencoder whose latent layer encodes cell-type fractions and
whose decoder weights collapse into a signature matrix. Its distinguishing
feature is an adaptive stage that refines the trained model directly on the
target bulk data, which lets it adjust to a tissue or condition the reference
does not perfectly match.

Chunks below are not evaluated during build. Epochs are reduced for
demonstration; use the function defaults for real analyses.

# Data

```{r data}
library(quasar)
library(QuasarDeconData)

sc_ref <- load_cov_imm("train")
pbulk  <- load_cov_pbulk(1)

bulk  <- pbulk$bulk_expression_profiles   # genes x samples
truth <- pbulk$ground_truth_proportions
```

# Simulate

```{r simulate}
tape_sim <- tape_simulate(
  sc_data      = sc_ref,
  celltype_col = "cell_type",
  samplenum    = 5000,
  n            = 500,
  sparse       = TRUE,
  random_state = 42
)
```

# Process

```{r process}
tape_proc <- tape_process(
  simudata           = tape_sim,
  real_bulk          = bulk,          # genes x samples
  variance_threshold = 0.98,
  scaler             = "mms"
)
```

`scaler = "mms"` is per-sample min-max; `"ss"` is per-sample standardisation
with a population standard deviation, matching sklearn rather than base R's
`sd()`.

# Train

```{r train}
tape_model <- tape_train(
  train_x    = tape_proc$train_x,
  train_y    = tape_proc$train_y,
  batch_size = 128,
  epochs     = 128,
  seed       = 0
)
```

The loss combines an L1 term on the latent proportions against the known
fractions with an L1 reconstruction term, so the decoder learns a signature
matrix at the same time as the encoder learns to deconvolve.

# Predict

```{r predict}
tape_res <- tape_predict(
  model      = tape_model,
  test_x     = tape_proc$test_x,
  genename   = tape_proc$genename,
  celltypes  = tape_proc$celltypes,
  samplename = tape_proc$samplename,
  adaptive   = TRUE,
  mode       = "overall"
)

head(tape_res$pred)
head(tape_res$sigm)
```

## Adaptive modes

The two modes differ substantially in cost and in what they return.

`"overall"` adapts one model jointly across all target samples and returns a
single signature matrix — one genes × cell-types profile for the whole dataset.

`"high-resolution"` adapts every sample with its own copy of the model,
returning a per-sample signature matrix for each cell type. Far more
informative, and far more expensive. `chunk_size` controls how many samples are
adapted concurrently as an ensemble of independent models; raise it to use more
GPU memory and less wall time.

```{r highres}
tape_hr <- tape_predict(
  model      = tape_model,
  test_x     = tape_proc$test_x,
  genename   = tape_proc$genename,
  celltypes  = tape_proc$celltypes,
  samplename = tape_proc$samplename,
  adaptive   = TRUE,
  mode       = "high-resolution",
  chunk_size = 16L
)

names(tape_hr$sigm)
dim(tape_hr$sigm[[1]])
```

Setting `adaptive = FALSE` skips refinement entirely and returns proportions
only, with `sigm` as `NULL`. This is much faster and worth using when the
reference already matches the target well.

# Wrapper

```{r wrapper}
tape_res <- tape(
  sc_data      = sc_ref,
  real_bulk    = bulk,
  celltype_col = "cell_type",
  samplenum    = 5000,
  n            = 500,
  epochs       = 128,
  adaptive     = TRUE,
  mode         = "overall"
)
```

# Evaluate

```{r evaluate}
quasar_prop_metrics(tape_res$pred, truth)$cell_type_rmse
```

# Reference

Chen, Y., Wang, Y., Chen, Y., Cheng, Y., Wei, Y., Li, Y., Wang, J., Wei, Y.,
Chan, T.-F., & Li, Y. (2022). Deep autoencoder for interpretable
tissue-adaptive deconvolution and cell-type-specific gene analysis. *Nature
Communications*, 13(1), 6735.

If you use this implementation, cite both the QUASAR paper and Chen et al.
